---
output: github_document
bibliography: '`r system.file("references.bib", package="vstar")`'
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# vstar

<!-- badges: start -->
<!-- badges: end -->

The goal of vstar is to test and estimate Vector Smooth Transition
Autoregression models (VSTAR).
The code is based on the specification from
[@terasvirta2014specification;@terasvirta2014linearity].

[@terasvirta2014specification] describes the most general model, allowing for
different transition variables and threshold values for every equation of the
system and every regime.
However, in the package implemented is much simpler, yet suitable specification,
which uses a common transition variable for all regimes and a common
threshold value in each regime.

## Installation

You can install the development version of vstar from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("vadim-zyamalov/vstar")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(vstar)
```

Let's generate an artifical data with two regimes using the following matrices:
\[A_1' = \left(\matrix{.5 & 0 \cr 0 & .5}\right)\]
\[A_2' = \left(\matrix{.8 & .1 \cr .1 & .8}\right)\]
The transition variable is a normal random variable with zero mean and unit
variance.
The threshold value is a median value of the transition variable.
The smoothness parameter \(\gamma\) equals 4.
We'll use a logistic transition function.

```{r data-generation, echo = FALSE}
N <- 1000
s <- rnorm(N)
thr_2reg <- median(s)
G.func <- vstar:::get.G.function("L")
sm <- s %x% t(vstar:::unity(1))
gm <- vstar:::unity(N) %x% t(4)
cm <- vstar:::unity(N) %x% t(thr_2reg)
G.mat_2reg <- G.func(sm, gm, cm)
A1 <- .5 * diag(2)
A2 <- matrix(c(.8, .1, .1, .8), ncol = 2)
B_2reg <- cbind(t(A1), t(A2))

y.2reg <- matrix(0, ncol = 2, nrow = 1001)
y.2reg[1, ] <- rnorm(2)
for (k in 1:N) {
    y.lag <- y.2reg[k, , drop = FALSE]
    PSI <- t(t(c(1, G.mat_2reg[k, ]))) %x% diag(2)
    y.2reg[k + 1, ] <- t(t(PSI) %*% t(B_2reg) %*% t(y.lag) + rnorm(2))
}
y.2reg <- y.2reg[52:(N + 1), 1:2, drop = FALSE]
colnames(y.2reg) <- c("y1", "y2")

s <- s[51:N]

plot.ts(y.2reg)
```

First, we prepare data for the LSTAR-model estimation:

```{r model-preparation}
model.2reg <- vstar.prepare(endo = c("y1", "y2"),
                            p = 1,
                            trans = s,
                            dataset = y.2reg)
```

The linearity test statistics reject the null of linearity:

```{r linearity-test}
round(linearity.test(model.2reg, J = 1), 4)
```

Next, we obtain the starting values for Nonlinear Least Squares by the grid
search.

```{r model-grid}
result.2reg <- vstar.grid(dataset = model.2reg,
                          m = 2,
                          gamma.limits = c(3, 5),
                          points = 200,
                          cores = 10)
```

Then we estimate the model with NLS.

```{r model-nls}
result.nls.2reg <- vstar.nls(result.2reg, tol = 1e-6, verbose = TRUE)
```

Then we print out the summary of the estimated model.

```{r model-summary}
summary(result.nls.2reg)
```

# References

<div id="refs"></div>
